---
title: 'Homework1 Abstract and Short Answers '
output:
  html_notebook: default
---

# Abstract

a. OBJECTIVE: An objective or description of the goal of the analysis

b. STUDY DESIGN: A brief description of the study design

c. METHODS: A methods section describing your statistical analysis (describe both the exploratory analysis and regression models with assumptions). Write this section so that we could replicate your analysis by reading your description.

d. RESULTS: A results section that includes a) descriptive statistics for the data (i.e. describe the sample in detail using summary statistics, be quantitative) b) a summary of your key findings including supporting numerical summaries (i.e. estimated slopes, confidence intervals, pvalues, etc.)

e. CONCLUSION: A conclusion specifically answering the objective of the study.


# Short Answer

## Question 1
\

R code:
```{r}
library(nlme)
library(here)
library(readr)
library(dplyr)
library(magrittr)

camp_primary <- read_csv(here("homework1", "camp_primary.csv"), col_names = T)
camp_primary %<>%
  filter(!is.na(POSFEV))

# change trt to a factor
camp_primary %<>% 
  mutate(trt = factor(trt, levels = c("0", "1", "2"),
                         labels = c("placebo", "budesonide", "nedocromil")))

# fit a random intercept model with an exponential variance matrix
## treat visitc as continuous, and assume no baseline difference for treatment groups

fit <- lme(POSFEV~visitc+trt:visitc,
          data=camp_primary,
          random=~1|id, 
          na.action=na.omit,
          correlation=corExp(form=~visitc|id))
summary(fit)

## F test for interaction term
anova(fit)
```

Interpret the coefficients for “visitc” and “1.trt#c.visitc” and provide statistical support for whether there is statistical evidence of a benefit of receiving budesonide compared to placebo for promoting long term improved pulmonary function among children with asthma.


**Model notations**

$$
E(Y_{ij}|X_{ij}) = \beta_0 + \beta_1visitc_{ij} + \beta_2visitc_{ij}*I(trt_i=budesonide) + 
                  \beta_3visitc_{ij}*I(trt_i=nedocromil); \\
                  i = 1,2,3,...,695; j=1,2,3,...,10
$$

- coefficients for “visitc”: 

$\beta_1$: the monthly rate of change in FEV1 in the placebo group in 48-months followup since randomization.

- coefficients for “1.trt#c.visitc”:

$\beta_2$: the difference in the monthly rate of change in FEV1 between the budesonide treatment and placebo group in 48-months followup since randomization 

- Model results: 

The F test for the Group * Time interaction terms is not statistically significant ($F=0.113,\hspace{1.5mm} p>.05$). It suggests there is no difference in the monthly rate of change in FEV1 between both treatment groups and the placebo group. So it will not substantially harm the fit of the model if we drop the Group * Time interaction terms.

The exponential + random intercept model results also show there is no statistically significant difference in the monthly rate of change in FEV1 between the budesonide treatment and placebo group in 48-months followup since randomization ($\beta2=-0.00022,\hspace{1.5mm} p>.05$).

Therefore, there is no statistical evidence of a benefit of receiving budesonide compared to placebo for promoting long term improved pulmonary function among children with asthma.
\


## Question 2

Using the fit of the model, estimate the correlation between the post-bronchodilator FEV1 at baseline assessment (visitc = 0) and 12-months post randomization (i.e. visitc = 12).

R code:

```{r}
fit <- lme(POSFEV~visitc+trt:visitc,
          data=camp_primary,
          random=~1|id, 
          na.action=na.omit,
          correlation=corExp(form=~visitc|id),
          method = "REML")
summary(fit)

## get variance matrix for the random effects
VarCorr(fit)

tau2 <-  as.numeric(VarCorr(fit)[1,1])
tau2 ## variance of the random intercept

sig2 <-  as.numeric(VarCorr(fit)[2,1])
sig2  ## variance of the residual

range <- coef(fit$modelStruct, unconstrained = FALSE)[2]
range

rho  <- exp(-1/range)
rho

## calculate correlation between the post-bronchodilator FEV1 at baseline assessment and 12-months post randomization

(tau2 + (rho^(12))*sig2)/(tau2 + sig2)
```

$$
\tau^2 = 0.355 \\
\sigma^2=0.0775 \\
\rho=e^{-1/range}=0.953 \\
range = 20.72129 \\
Corr(Y_{i0},Y_{i12})=\frac{\tau^2+\rho^{12}\sigma^2}{\tau^2+\sigma^2} = 0.921
$$

So, the correlation between the post-bronchodilator FEV1 at baseline assessment (visitc = 0) and 12-months post randomization is:

$Corr(Y_{i0},Y_{i12})=0.921$.
\


## Question 3 

Describe an approach you could take to assess how well the model you fit describes the within subject correlation in the CAMP study.

We can compare the estimated correlation for observations u-lag units apart with the correlation we get from the empiral correlation matrix. Also, we can compare it with the correaltion from the auto-correlation function. For example. the correlation for observations 12-lag units apart is 0.921 above. The empirical correlation between baseline and 12-months we got fro the exploratory data analysis is 0.955. So it appears the model we fit describes the within subject correlation in the CAMP study pretty well.
\


## Question 4 

1. Model notations and definitions for all variables and coefficients

$$
E(Y_{ij}|X_{ij}) = \beta_0 + \beta_1visitc_{ij} + \beta_2visitc_{ij}*I(trt_i=budesonide) + 
                  \beta_3visitc_{ij}*I(trt_i=nedocromil) + \\
                  \beta_4(visitc_{ij}>52)*(visitc_{ij}-52) + \\
                  \beta_5(visitc_{ij}>52)*(visitc_{ij}-52)*I(trt_i=budesonide) + \\
                  \beta_6(visitc_{ij}>52)*(visitc_{ij}-52)*I(trt_i=nedocromil) 
                  \\ i = 1,2,3,...,695; j=1,2,3,...,16
$$


**Definitions**

$trt_{ij}$: treatment groups: placebo; budesonide; nedocromil

$visitc_{ij}$: 16 time responses across 108 months since randomization: 0, 2, 4, 12, 16, 24, 28, 36, 40, 48, 52, 60, 72, 84, 96, 108

$visitc_{ij}>52$: 

- 0, if follow-up months fewer than or equal 52 months;

- 1, if follow-up months greater than 52 months.

$visitc_{ij}-52$: the numbers of months after 52 months.


$\beta_1$: monthly rate of change in FEV1 in the placebo group before 52 months since randomization.

$\beta_2$: the difference in the monthly rate of change in FEV1 between the budesonide treatment and placebo group before 52 months since randomization.

$\beta_3$: the difference in the monthly rate of change in FEV1 between the nedocromil treatment and placebo group before 52 months since randomization.

$\beta_4$: the difference in monthly rate of change in FEV1 in the placebo group after vs. before 52 months since randomization.

$\beta_5$: the difference in the monthly rate of change in FEV1 between the budesonide treatment and placebo group after vs. before 52 months since randomization.

$\beta_6$: the difference in the monthly rate of change in FEV1 between the nedocromil treatment and placebo group after vs. before 52 months since randomization.


2. provide the three relevant hypothesis tests 

$$
H_0 : \beta_4=\beta_5=\beta_6=0 \\
H_a : \beta_4=\beta_5=\beta_6 \ne 0 \\
$$


3. using the same correlation model as above (exponential), fit the model and conduct the three hypothesis tests


```{r}
## read continuation file
camp_cont <- read_csv(here("homework1", "camp_continuation.csv"), col_names = T)

## remove missingness
camp_cont %<>%
  filter(!is.na(POSFEV))

## change trt to factor
camp_cont %<>% 
  mutate(trt = factor(trt, levels = c("0", "1", "2"),
                         labels = c("placebo", "budesonide", "nedocromil")))

## generate a spline term at 52 months
camp_cont %<>% 
  mutate(visitc_52 = visitc-52, 
       visitc_sp = (visitc_52>0)*visitc_52,
       visitc_52_trt = visitc_52*(as.numeric(trt)-1),
       visitc_sp_trt = visitc_sp*(as.numeric(trt)-1))
```


```{r}
## fit a exponential model with a random intercept and a spline term at 52 months

fit_sp <- lme(POSFEV ~ visitc + trt:visitc + visitc_sp + visitc_sp:trt,
          data=camp_cont,
          random=~1|id, 
          na.action=na.omit,
          correlation=corExp(form=~visitc|id))

summary(fit_sp)

anova(fit_sp)
```


4. 

The F tests for the placebo and treatment spline terms are statistically significant ($F_{placebo}=30.77,\hspace{1.5mm} p<.0001; \hspace{2mm} F_{trt}=4.16,\hspace{1.5mm} p<.0001$). It suggests there is statistically significant difference in the monthly rate of change in FEV1 during the continuation study and the primary study in the treatment groups.

The exponential + random intercept model results also show there is statistically significant difference in the monthly rate of change in FEV1 in the placebo group after vs. before 52 months since randomization ($\beta_3=-0.005,\hspace{1.5mm} p<.0001$).

Besides, there is statistically significant difference in the monthly rate of change in FEV1 between the nedocromil treatment and placebo group after vs. before 52 months since randomization ($\beta_6=0.004,\hspace{1.5mm} p<.005$).

However, there is no statistically significant difference in the monthly rate of change in FEV1 between the budesonide treatment and placebo group after vs. before 52 months since randomization ($\beta_5=0.001,\hspace{1.5mm} p>.05$).

Therefore, the monthly rate of change in FEV1 during the continuation study is the same as the monthly rate of change in FEV1 during the primary study for the budesonide treatment group, but the nedocromil treatment and placebo group have different monthly rate of change in FEV1 during the continuation study comparing with the primary study.
\

